Augmentation of a Historical Harpsichord Keyboard Replica for Haptic-Enabled Interaction in Museum Exhibitions

### 2025-04-09 src/article.tex
address part of #45

However, the delicate mechanisms of these instruments and their historical significance mean they are only played under strict conditions—by experienced historical keyboard performers or young musicians under strict supervision from the curator. 
To enhance accessibility and engagement, the museum commissioned the replica (Figure \ref{fig:teaser}) of a historical keyboard built in the tradition of the Italian school, which is the subject of this paper.
The keyboard, which museum visitors can use without special permission, presents two choirs of muted strings and can generate MIDI messages. The interface is presently linked to a commercial software sampler; however, the ultimate aim is to make the sound of instruments in the collection that can no longer be maintained in playable condition accessible. Museum visitors are invited to play the interface and listen through a pair of headphones. This work outlines the technological aspects of the interface's construction and offers reflections on its role within the \anon{Tagliavini} Collection and potential application within the wider musical instrument museum context.

@@

The final design, shown in Figures \ref{fig:teaser} and \ref{fig:details}, is a 49-key, two-register harpsichord keyboard replica. Since the Tagliavini collection hosts primarily early plucked stringed keyboards of Italian origin, the keyboard layout was deliberately modelled after early Italian harpsichords, leveraging the human tendency to be influenced by visual elements when making musical judgments \cite{Tsay2013}. This effect is particularly relevant in musical instruments, as demonstrated by studies conducted by Fritz \emph{et al.} \cite{Fritz2012, Fritz2014, Fritz2017}. Instead of attempting to mitigate this influence, the design embraces it. This phenomenon, akin to a `musical instrument McGurk effect’, is one of the reasons why the electronics have been concealed from view. The aesthetics of the interface enhance its likelihood of being perceived as an `authentic' musical instrument. Given the extensive research into modelling and recreating piano action \cite{Cadoz1990, Gillespie1996, Timmermans2020}, the visual component may provide the final persuasive element necessary for acceptance, similar to the way visual perception affects judgments of musical performance \cite{Tsay2013}.

\begin{figure}[!b]  
  \centering
  \includegraphics[width=\linewidth,trim={0 0 0 0.5cm},clip]{src/images/tagging-jacks-3.jpg} 
  \caption{Gradient stickers applied to the side of the jack body. The coarse gradient scale was selected to maximise signal excursion while preserving signal readout stability.}
  \Description{Close of two jacks with gradient stickers applied to each. Both jack lie on top of the vinyl sheet that was used for printing. The gradient is a long black bar that terminates in 7 rectangles of grey over an 8mm span. The remainder of the sticker is white. The jacks have the numbers 46 and 49 written on them at different offset illustrating the numbering system to identify their position along the keyboard.} 
  \label{fig:jack-tags}
\end{figure}

@@

\begin{figure}
  \centering
  \includegraphics[width=\linewidth]{src/images/49-key-bottom-sensors-no-keys.jpg} 
  \caption{Underside of the full model keyboard, showing two chambers: the front chamber (top) and the rear chamber (bottom).} 
  \Description{Underside view of the models with keys removed. A cross brace split the case into four quadrants the top two are the front chambe and the bottom are the rea chamber. Visible inside are the PCB which are connected to each other with multicoloured ribbon cable. From the right 2 long cables, one for sensor signals the other for LED and multiplexer signals. Both sets of cables terinate at the controller board. The PCBs are attached with white plastic supports to the underside of the wrest plank, a thick block of wood in which the tuning pegs are screwed.} 
  \label{fig:49-key-bottom}
\end{figure}

@@

\begin{figure*}
    \centering
    \includegraphics[width=\linewidth]{src/images/block-diagram-2.pdf}
    \caption{Block diagram of PCB connections. A separate sensor signal is routed to the Arduino. LED and multiplexer (MUX) controls signal are daisy-chained through each sensor PCB.}
    \Description{The block diagram shows seven blocks each with sensor board written on them. Each sensor block has a solid arrow which connects to a block labelled Arduino Nano 33 BLE. A dashed arrow is drawn from the Ardnuino Nano 33 BLE box and travels through each sensors board. A legend shows th soldi line represnts the flow of the sensor signal and the dashed arrow the flow of LED an multiplexer control signals. Angled arrows from the sensor board point towards and away from a hatched lined representing the surface of the jacks, which illustartes the reflection of light. Lines from boxes labelled FRAM and Rotary Encoder point to and from the box labelled Arduino Nano 33 BLE.}
    \label{fig:system-block-diagram}
\end{figure*}

@@

\begin{figure}[!b] 
  \centering
  \includegraphics[width=\linewidth,trim={0 1.5cm 0 1.5cm},clip]{src/images/sensor-board-w-baffles.jpeg}
    \\
  \includegraphics[width=\linewidth]{src/images/sensor-board-reverse-side-anon.png} 
  \caption{Sensor board showing the sensor side (top) and rear side with RGB LEDs and terminal connections (bottom). The \texttt{SIG} terminal (Top Left) is routed to an ADC channel. Power \texttt{5V, 3V, GND}, multiplexer channels \texttt{A, B, C} and LED PWM input \texttt{DI} and output \texttt{DO}  connect from the previous sensor board (Left) and are routed to the next (Right).}
  \Description{Two images one of the front and one of the rear of a single sensor board. The top image shows the baffles in place and how they enclose the field of view of each sensor. The bottom image shows the wires from neighbouring circuit boards terminating and the line of 7 RGB LEDs in the same spacing as the LEDs.} 
  \label{fig:sensor-reverse}
\end{figure}

@@

This work has incorporated and developed further the ideas of reusing and complimenting the `old' in digital musical instrument design as put forward in the article `The \emph{O} in NIME' \cite{Masu_NIME_2023}. By integrating historical keyboard-building traditions with digital augmentation, it offers a possibility to extend the practical lifespan of musical heritage while maintaining the tactile qualities of the original instrument. Rather than prioritising technical novelty, the project demonstrates how digital interventions can support meaningful interaction with a shared musical heritage, ensuring its continued relevance in contemporary museum contexts.

>>>>

However, the delicate mechanisms of these instruments and their historical significance mean they are only played under strict conditions—by experienced historical keyboard performers or young musicians under strict supervision from the curator. 
To enhance accessibility and engagement, the museum commissioned the replica (Figure \ref{fig:teaser}) of a historical keyboard built in the tradition of the Italian school, which is the subject of this paper.
% target audience: The target audience for the interface is visitors to the museum, who would interact with instrument if it were working order.
% The keyboard, which museum visitors can use without special permission, presents two choirs of muted strings and can generate MIDI messages. 
The keyboard targets all visitors to the museum, regardless of playing ability, and can be used without special permission.
Jacks that pluck two choirs of muted strings, across 49-keys, are used to generate MIDI messages that are sent to a connected computer for audio synthesis. 
The interface is presently linked to a commercial software sampler; however, the ultimate aim is to make the sound of instruments in the collection that can no longer be maintained in playable condition accessible. Museum visitors are invited to play the interface and listen through a pair of headphones. This work outlines the technological aspects of the interface's construction and offers reflections on its role within the \anon{Tagliavini} Collection and potential application within the wider musical instrument museum context.

@@

The final design, shown in Figures \ref{fig:teaser} and \ref{fig:details}, is a 49-key, two-register harpsichord keyboard replica. Since the \anon{Tagliavini collection} hosts primarily early plucked stringed keyboards of Italian origin, the keyboard layout was deliberately modelled after early Italian harpsichords, leveraging the human tendency to be influenced by visual elements when making musical judgments \cite{Tsay2013}. This effect is particularly relevant in musical instruments, as demonstrated by studies conducted by Fritz \emph{et al.} \cite{Fritz2012, Fritz2014, Fritz2017}. Instead of attempting to mitigate this influence, the design embraces it. This phenomenon, akin to a `musical instrument McGurk effect’, is one of the reasons why the electronics have been concealed from view. The aesthetics of the interface enhance its likelihood of being perceived as an `authentic' musical instrument. Given the extensive research into modelling and recreating piano action \cite{Cadoz1990, Gillespie1996, Timmermans2020}, the visual component may provide the final persuasive element necessary for acceptance, similar to the way visual perception affects judgments of musical performance \cite{Tsay2013}.

\begin{figure}[!b]  
  \centering
  \includegraphics[width=\linewidth,trim={0 0 0 0.5cm},clip]{src/images/tagging-jacks-3.jpg} 
  \caption{Gradient stickers applied to the side of the jack body. The coarse gradient scale was selected to maximise signal excursion while preserving signal readout stability.}
  \Description{Close of two jacks with gradient stickers applied to each. Both jack lie on top of the vinyl sheet that was used for printing. The gradient is a long black bar that terminates in 7 rectangles of grey over an 8 millimetre span. The remainder of the sticker is white. The jacks have the numbers 46 and 49 written on them at different offset illustrating the numbering system to identify their position along the keyboard.} 
  \label{fig:jack-tags}
\end{figure}

@@

\begin{figure}
  \centering
  \includegraphics[width=\linewidth]{src/images/49-key-bottom-sensors-no-keys.jpg} 
  \caption{Underside of the full model keyboard, showing two chambers: the front chamber (top) and the rear chamber (bottom).} 
  \Description{Underside view of the models with keys removed. A cross brace split the case into four quadrants the top two are the front chamber and the bottom are the rear chamber. Visible inside are the PCB which are connected to each other with multicoloured ribbon cable. From the right 2 long cables, one for sensor signals the other for LED and multiplexer signals. Both sets of cables terminate at the controller board. The PCBs are attached with white plastic supports to the underside of the wrest plank, a thick block of wood in which the tuning pegs are screwed.} 
  \label{fig:49-key-bottom}
\end{figure}

@@

\begin{figure*}
    \centering
    \includegraphics[width=\linewidth]{src/images/block-diagram-2.pdf}
    \caption{Block diagram of PCB connections. A separate sensor signal is routed to the Arduino. LED and multiplexer (MUX) controls signal are daisy-chained through each sensor PCB.}
    \Description{The block diagram shows seven blocks each with sensor board written on them. Each sensor block has a solid arrow which connects to a block labelled Arduino Nano 33 BLE. A dashed arrow is drawn from the Ardnuino Nano 33 BLE box and travels through each sensors board. A legend shows the solid line represents the flow of the sensor signal and the dashed arrow the flow of LED an multiplexer control signals. Angled arrows from the sensor board point towards and away from a hatched lined representing the surface of the jacks, which illustrates the reflection of light. Lines from boxes labelled FRAM and Rotary Encoder point to and from the box labelled Arduino Nano 33 BLE.}
    \label{fig:system-block-diagram}
\end{figure*}

@@

\begin{figure}[!b] 
  \centering
  \includegraphics[width=\linewidth,trim={0 1.5cm 0 1.5cm},clip]{src/images/sensor-board-w-baffles.jpeg}
    \\
  \includegraphics[width=\linewidth,trim={4.45cm 4.5cm 5.25cm 4cm},clip]{src/images/sensor-board-reverse-side.png} 
  \caption{Sensor board showing the sensor side (top) and rear side with RGB LEDs and terminal connections (bottom). The \texttt{SIG} terminal (Top Left) is routed to an ADC channel. Power \texttt{5V, 3V, GND}, multiplexer channels \texttt{A, B, C} and LED PWM input \texttt{DI} and output \texttt{DO}  connect from the previous sensor board (Left) and are routed to the next (Right).}
  \Description{Two images one of the front and one of the rear of a single sensor board. The top image shows the baffles in place and how they enclose the field of view of each sensor. The bottom image shows the wires from neighbouring circuit boards terminating and the line of 7 RGB LEDs in the same spacing as the LEDs.} 
  \label{fig:sensor-reverse}
\end{figure}

@@

This work has incorporated and developed further the ideas of reusing and complimenting the `old' in digital musical instrument design as put forward in the article `The \emph{O} in NIME' \cite{Masu_NIME_2023}. By integrating historical keyboard-building traditions with digital augmentation, it offers a possibility to extend the practical lifespan of musical heritage while maintaining the tactile qualities of the original instrument. Rather than prioritising technical novelty, the project demonstrates how digital interventions can support multi-modal interaction with a shared musical heritage, ensuring its continued relevance in contemporary museum contexts.
